Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-04-21 demystify.tex
updates

Many students get letters from supervisors from companies where they did internships or are working.  It is OK as long as it is a research-based personalized letter. Again, the emphasis here is \emph{research}, i.e., the letters should describe your research experiences and potential. Letters focusing on your course work or non-research projects at a company won't carry much weight.

@@

As mentioned, if you're admitted to a \emph{good} CS PhD program, you should not have to worry about funding!
In the US, the common types of funding for PhDs are \emph{graduate teaching assistant} (GTA or TA), \emph{graduate research assistant} (GRA or RA), and \emph{Fellowship}.
RA is paid by a prof. for you to do their research. TA is paid by the department for you to help with teaching. Finally, fellowship is an independent funding that can come from a school, a company, or an organization. \autoref{tab:funding} summarizes the differences.
Note that funding is typically more available for PhD students than
Masters.

@@

Typically, it takes 5--7 years for CS PhD in the US.  This can be longer than CS PhDs in other countries, which might require MS first (recall that CS PhD programs in the US do not require MS and you can get MS along the way to PhD). Within these 5--7 years, CS PhD students often take a ``leave of absence'' for 1--2 semesters to do internship at companies and research labs.

The first 2 years you typically take coursework (somewhat equivalent to an MS study), find an adviser, learn how to do research.  The next 2--3 years you focus on your research, form dissertation topic, and get results published. The last 1-2 years you continue to publish, write and defend your dissertation, and look for job.
In many cases you might take a summer or two off to do an internship to get additional research opportunities.
The PhDComics figure on top shows the ``ambition'' level of a PhD student over their years of study (they miss the 6--7th year when the ambition is \emph{``Just let me graduate''}).

@@

\emph{Course work}: as mentioned above in the US you will spend the first couple of years taking classes and exploring potential adviser and research topics. 
After that, you also have to pass a series of exams during your PhD, e.g., qualifying exam, comprehensive exam, thesis proposal defense\footnote{The word ``ABD'' (all but dissertation) is used in the US to refer to a PhD candidate who have finished all course work and exams and only need to write and defend their dissertation.}.

@@

PhD students often ask why their salary is so low compared to ludicrous grants their advisers get. They also wonder why their offer letters sometimes show that their benefits are higher than what they receive (i.e., stipend).

@@

\section{Will I be miserable during my a PhD?}\label{sec:happy}
There are many stories of how students are mistreated, stressed, and miserable. Issues including bad relationships with professors, conflicts with co-authors and lab mates, feeling discriminated against (e.g., because you're an international student), \emph{do} exist, and it is good to be aware of those.  However, in reality, there are many good mentors, supportive lab mates and department, and so on.  So don't let social media make you feel pessimistic and deter your quest to advance knowledge.

@@

\paragraph{Funding \autoref{sec:funding}} You have more opportunities for funding, e.g., through government scholarships for US citizens and permanent residents.  You can also apply to specific programs \emph{before} you start your PhD, e.g., NSF Graduate Research Fellowship Program (GRFP) and Hertz Foundation Fellowship.  These fellowships are very competitive and can significantly improve your admission chance if you get them.

>>>>

Many students get letters from supervisors from companies where they did internships or are working.  It is OK as long as it is a research-based personalized letter. Again, the emphasis here is \emph{research}, i.e., the letters should describe your research experiences and potential. Letters focusing on your coursework or non-research projects at a company won't carry much weight.

@@

As mentioned, if you're admitted to a \emph{good} CS PhD program, you should not have to worry about funding!
In the US, the common types of funding for PhDs are \emph{graduate teaching assistant} (GTA or TA), \emph{graduate research assistant} (GRA or RA), and \emph{Fellowship}.
RA is paid by a prof. for you to do their research. TA is paid by the department for you to help with teaching. Finally, fellowship is independent funding that can come from a school, a company, or an organization. \autoref{tab:funding} summarizes the differences.
Note that funding is typically more available for PhD students than
Masters.

@@

Typically, it takes 5--7 years for CS PhD in the US.  This can be longer than CS PhDs in other countries, which might require MS first (recall that CS PhD programs in the US do not require MS and you can get MS along the way to PhD). Within these 5--7 years, CS PhD students often take a ``leave of absence'' for 1--2 semesters to do internships at companies and research labs.

The first 2 years you typically take coursework (somewhat equivalent to an MS study), find an adviser, and learn how to do research.  The next 2--3 years you focus on your research, form a dissertation topic, and get results published. The last 1-2 years you continue to publish, write and defend your dissertation, and look for a job.
In many cases you might take a summer or two off to do an internship to get additional research opportunities.
The PhDComics figure on top shows the ``ambition'' level of a PhD student over their years of study (they miss the 6--7th year when the ambition is \emph{``Just let me graduate''}).

@@

\emph{Course work}: as mentioned above in the US you will spend the first couple of years taking classes and exploring potential adviser and research topics. 
After that, you also have to pass a series of exams during your PhD, e.g., qualifying exam, comprehensive exam, thesis proposal defense\footnote{The word ``ABD'' (all but dissertation) is used in the US to refer to a PhD candidate who has finished all course work and exams and only needs to write and defend their dissertation.}.

@@

PhD students often ask why their salary is so low compared to the large grants their advisers get. They also wonder why their offer letters sometimes show that their benefits are higher than what they receive (i.e., stipend).

@@

\section{Will I be miserable during my a PhD?}\label{sec:happy}
There are many stories of how students are mistreated, stressed, and miserable. Issues including bad relationships with professors, conflicts with co-authors and lab mates, feeling discriminated against (e.g., because you're an international student), \emph{do} exist, and it is good to be aware of those.  However, in reality, there are many good mentors, supportive lab mates and departments, and so on.  So don't let social media make you feel pessimistic and deter your quest to advance knowledge.

@@

\paragraph{Funding \autoref{sec:funding}} You have more opportunities for funding, e.g., through government scholarships for US citizens and permanent residents.  You can also apply to specific programs \emph{before} you start your PhD, e.g., NSF Graduate Research Fellowship Program (GRFP) and Hertz Foundation Fellowship.  These fellowships are very competitive and can significantly improve your admission chances if you get them.
